\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\geometry{margin=1in}

\title{Resolving Cosmological Tension via Minimal Kinetic Coupling (MKC) and Spectral Rigidity}
\author{Inacio F. Vasquez}
\date{2026}

\begin{document}
\maketitle

 Technical Manuscript Outline: DFM-MKC-Draft
Title: Resolving Cosmological Tension via Minimal Kinetic Coupling (MKC) and Spectral Rigidity
Registry ID: EXT-COS-01
Classification: Formal Closure / Information Topology
1. Introduction: The Topology of Tension
The Impasse.
The observed discrepancies in the H₀ (expansion rate) and S₈ (structure growth) parameters are identified not as calibration artifacts, but as structural failures of the ΛCDM assumption that dark matter and dark energy form dynamically disjoint sectors.
The Proposition.
Introduction of the Dark Fluid Model (DFM), in which dark matter and dark energy arise as phases of a single information-bearing manifold with continuous kinetic structure.
2. Theoretical Framework: Minimal Kinetic Coupling (MKC)
The Action.
Modification of the Einstein–Hilbert action by the inclusion of a minimal kinetic coupling term
ξ
 
(
∂
ϕ
)
2
,
ξ(∂ϕ) 
2
 ,
linking the dark sector scalar field to spacetime dynamics.
The Coupling Invariant.
The parameter 
ξ
ξ is established as a structural invariant preventing topological tearing during the transition from matter-dominated to energy-dominated cosmological eras.
Spectral Gap Requirement.
Application of the URF rigidity condition
λ
1
 ⁣
(
Δ
M
K
C
)
>
0
,
λ 
1
​	
 (Δ 
MKC
​	
 )>0,
ensuring that global expansion remains rigid whenever local kinetic homogeneity is preserved.
3. Resolution of the H₀ Parameter
Mechanism.
The MKC term induces a smooth interpolation in the effective equation of state 
w
e
f
f
w 
eff
​	
 , reconciling early-time CMB constraints with late-time distance-ladder measurements.
Deterministic Fit.
Derivation of a URF-consistent expansion rate
H
0
≈
72.9
 
k
m
 
s
−
1
 
M
p
c
−
1
,
H 
0
​	
 ≈72.9 kms 
−1
 Mpc 
−1
 ,
in contrast to the mutually inconsistent values produced by disjoint-sector models.
4. Resolution of the S₈ Growth Parameter
Stabilized Growth.
Structure-growth suppression in ΛCDM is reinterpreted as an artifact of neglecting kinetic coupling.
By analogy with the Biological Friction Framework, MKC stabilizes gravitational potentials without invoking ad hoc damping, yielding the observed S₈ values.
5. Deterministic Auditing (URF-SG Standard)
Machine Verifiability.
Formal verification via executable witnesses (topology_check.json, mkc_solver.py) demonstrating compliance with URF-SG requirements.
Logical Width Dependency.
Proof that for 
k
≥
4
k≥4, local dark-fluid homogeneity in FOᵏ logic forces the globally observed late-time expansion rate.
6. Conclusion: Remaining External Uncertainty
Closure Statement.
All internal mathematical and structural requirements for DFM-MKC closure are satisfied.
Roadmap.
Outstanding uncertainty is external: validation, citation, and institutional adoption rather than theoretical consistency.

\end{document}
